---
title: "methcp: User’s Guide"
author: "Boying Gong"
date: "`r Sys.Date()`"
output: 
      html_document:
        toc: true
        toc_float: true
        number_sections: true
        toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Load packages. 

```{r, message=FALSE}
library(bsseq)
library(methcp)
```

# Two-group comparison

In this section, we use the CpG methylation data from an Arabidopsis dataset available from GEO with accession number [GSM954584](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM954584). We take a subset of chromosome 1 and 2 from each of the samples and perform differential analysis between the wild-type plants and the H2A.Z mutant plants, which we refer to as `treatment` and `control` in the rest of the document. 

## Read data

We use a well-developed Bioconductor package `bsseq` to load the store the raw data. Below is an example of how to read raw counts using `bsseq`. For more operations regarding the `bsseq` object, please refer to their [User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/bsseq/inst/doc/bsseq.html). 

```{r readData}
sample_names <- c(paste0("control", 1:3),
                  paste0("treatment", 1:3))
n_sample <- length(sample_names)

raw_files <- system.file(
  "extdata", paste0(sample_names, ".txt"), package = "methcp")

bs_object_list <- list()
for (i in 1:n_sample){
  dt <- read.table(raw_files[i], stringsAsFactors = FALSE, header = TRUE)
  bs_object_list[[i]] <- BSseq(chr = dt$Chr,
                               pos = dt$Pos,
                               M = as.matrix(dt$M),
                               Cov = as.matrix(dt$Cov),
                               sampleNames = sample_names[i])
}

bs_object <- do.call("combine", c(bs_object_list))
```

The `bsseq` object.

```{r showBSobject}
bs_object
```

Header of the raw file for one of the samples.

```{r}
head(dt)
```

## Calculate the statistics

We calculate the per-cytosine statistics using two different test `DSS` and `methylKit` using the function `calcLociStat`. This function returns a `MethCP` object that is used for the future segmentation step. We allow parallelized computing when there are multiple chromosomes in the dataset. Note that the number of cores used as indicated in the parameter `mc.cores` should not exceeds the number of chromosomes in the dataset.

```{r calcStat}
group1 <- paste0("control", 1:3)
group2 <- paste0("treatment", 1:3)

obj_DSS <- calcLociStat(bs_object, group1, group2, test = "DSS", mc.cores = 1)
obj_methylKit <- calcLociStat(bs_object, group1, group2, test = "methylKit", mc.cores = 1)
```

```{r obj_DSS}
obj_DSS
```

```{r obj_methylKit}
obj_methylKit
```

In cases of experiments other than two-group comparison, we use the calculated statistics and create a `MethCP` object.

```{r createmethcp}
obj <- methcpFromStat(chr = c("Chr01", "Chr01", "Chr02", "Chr02"), 
                      pos = c(12, 109, 42, 78), 
                      pvals = c(0, 0.5, 0.13, 0.01), 
                      effect.size = c(0.76, -0.2, 0.31, 0.66), 
                      testName = "UserDefinedTest")
```

```{r}
obj
```

## Segmentation

`segmentMethCP` performs segmentation on a `MethCP` object. We allow parallelized computing when there are multiple chromosomes in the dataset. Different from `calcLociStat` function in the previous section, we do not put any constraint on the number of cores used. Please see the documentation for adjusting the parameters used in the segmentation.

```{r segmentation}
obj_DSS <- segmentMethCP(obj_DSS, bs_object, region.test = "weighted-coverage", mc.cores = 1)
obj_methylKit <- segmentMethCP(obj_methylKit, bs_object, region.test = "fisher", mc.cores = 1)
```

```{r}
obj_DSS
```

```{r}
obj_methylKit
```

## Significant regions

Use function `getSigRegion` on a `MethCP` object to get the list of DMRs.

```{r}
region_DSS <- getSigRegion(obj_DSS)
head(region_DSS)
```

```{r}
region_methylKit <- getSigRegion(obj_methylKit)
head(region_methylKit)
```

# A time-course example

MethCP is flexible for a wide variety of experimental designs. We apply MethCP on an Arabidopsis thaliana seed germination dataset available from the GEO with accession number [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE94712](GSE94712) The data is generated from two replicates of dry seed and germinating seeds of wild-type plants (Col-0) and ros1 dml2 dml3 (rdd) triple demethylase mutant plants at 0-4 days after imbibition for 4 days (DAI). For cytosine-based statistics, we fit linear models on the methylation ratios and test the differences between the time coefficient of Col-0 and rdd.

Read the meta data.

```{r}
meta_file <- system.file(
  "extdata", "meta_data.txt", package = "methcp")
meta <- read.table(meta_file, sep = "\t", header = TRUE)
```

Read the counts data.

```{r}
raw_files <- system.file(
  "extdata", paste0(meta$SampleName, ".tsv"), package = "methcp")

bs_object_list <- list()
for (i in 1:length(raw_files)){
  dt <- read.table(raw_files[i], stringsAsFactors = FALSE, header = TRUE)
  bs_object_list[[i]] <- BSseq(chr = dt$chr,
                               pos = dt$pos,
                               M = as.matrix(dt$M),
                               Cov = as.matrix(dt$Cov),
                               sampleNames = meta$SampleName[i])
}

bs_object <- do.call("combine", c(bs_object_list))
bs_object
```

Apply coverage filter.

```{r}
groups <- split(1:nrow(meta), meta$Condition)
coverages <- as.data.frame(getCoverage(bs_object, type = "Cov"))
filter <- rowSums(coverages[, meta$SampleName[groups[[1]]]] != 0) >= 3 &
  rowSums(coverages[, meta$SampleName[groups[[2]]]] != 0) >= 3
bs_object <- bs_object[filter, ]
```

Calculate the statistics. A dataframe of the meta data will be passed to function `calcLociStatTimeCourse`. Note that there must be columns named `Condition`, `Time` and `SampleName` in the dataframe.

```{r}
obj <- calcLociStatTimeCourse(bs_object, meta)
```

```{r}
obj
```

Segmentation.

```{r}
obj <- segmentMethCP(obj, bs_object, region.test = "stouffer", mc.cores = 1)
```

```{r}
head(obj@segmentation)
```


Get the DMRs.

```{r}
regions <- getSigRegion(obj)
```

```{r}
head(regions)
```


# Session info

```{r}
sessionInfo() 
```
